### 2025-07-27 response.tex

\documentclass[12pt,a4paper,preprint]{article}

\usepackage{setspace}
\usepackage{xcolor}
\usepackage{amssymb}
\usepackage{softdev}
\usepackage{quoting}
\usepackage{xspace}
\usepackage[many]{tcolorbox}
\newtcolorbox{blockquote}{%
  sharp corners,
  colback=gray!5,
  colframe=gray!60,
  leftrule=2pt,
  enhanced,
  before skip=6pt, after skip=18pt,
  boxrule=0pt,
  left=6pt, right=4pt,
  borderline west={2pt}{0pt}{gray!60},
  fontupper=\scriptsize\sffamily\selectfont
}

\setlength{\parindent}{0pt}
\setlength{\parskip}{2.0ex plus0.5ex minus0.2ex}

%\newcommand\laurie[1]{\mynote{Laurie}{#1}}
%\newcommand\jake[1]{\mynote{Jake}{#1}}

\newcommand\Egcrc{E$_\textrm{gcvs}$\xspace}
\newcommand\Eelision{E$_\textrm{elision}$\xspace}
\newcommand\Epremopt{E$_\textrm{premopt}$\xspace}

\begin{document}

\date{}  % This removes the date
\title{Response to Reviewer Feedback}
\maketitle

We thank the reviewers for their thoughtful reviews! While our \texttt{revisions R2} submission contains a PDF diff that shows prose modifications, we recognize that many review comments centered on our evaluation approach. Since \texttt{latexdiff} cannot effectively display changes to figures and tables, this document serves as a brief guide to our major revisions.

Below, we highlight the most significant changes made in response to reviewer input, including references to new visualizations that directly address the points raised during review.

\subsection*{Better Justification around Finalizers}

\begin{blockquote}
explicitly discuss and present the issues around finalizations semantics, as learned from other languages...\\

... This should take the form of a) explicit acknowledgement of the issue of this as a solution (as per the JEP) b) explain why such an undesirable solution is unavoidable (if it is) c) clearly explain how the problems with finalization semantics relate to destructor semantics which Rust has already bought into (i.e. exactly how much introducing this approach makes things worse or not).
\end{blockquote}

We have updated Section 4.1 to include a more detailed discussion of how finalization semantics relate to destructor semantics in Rust in particular. We offer a concrete example of an instance where finalizers can actually lead to more drop methods being run than their equivalent destructor based counterparts -- across panic unwind boundaries on different mutator threads.

We also update Section 4.2 to more explicitly address finalizer challenges (such as latency) as they relate to JEP-421.

In Section 4 we mention that despite the issues surrounding finalization, Rust's heavy reliance on destructor-based code, and Alloy's desire allow GC interaction with these objects, finalization is inevitable.

\subsection*{Performance Evaluation}

\begin{blockquote}
include the non-BDW baseline in the evaluation and discuss the utility of this approach w.r.t. that performance result. Ideally this would include a discussion of how much of the performance deficit is likely redeemable via "engineering", and how much is likely to be intrinsic (i.e. due to unavoidable conservatism), so that we can get a better intuition of the plausible "best case" for this approach.
\end{blockquote}

We have standardized all visualizations in Section 8 by normalizing against
appropriate baselines. Each experiment uses a specific baseline for comparison:
non-BDW allocator benchmarks for the \Egcrc experiment (RC comparison)
\jake{TODO link to figures}, naive Alloy without elision for the \Eelision
experiment \jake{TODO link to figures}, and barrier-free execution for the
\Epremopt experiment \jake{TODO link to figures}. All results are presented as
ratio relative to these baselines rather than raw values. For the interested
reader, we provide the raw values for each experiment in appendix tables.

We offer a discussion of the engineering cost in the prose in section \jake{TODO: section}

\subsection*{Conclusion}

\begin{blockquote}
add a discussion thats sums up the overall benefits/drawbacks of this particular approach, possibly depending on specific scenarios/use cases, and perhaps outlines directions for future work from a conceptional perspective, i.e., possible design directions that could be investigated
\end{blockquote}

\end{document}

### 2025-07-27 response.tex

\newcommand\Egcrc{E$_\textrm{gcvs}$\xspace}
\newcommand\Eelision{E$_\textrm{elision}$\xspace}
\newcommand\Epremopt{E$_\textrm{premopt}$\xspace}
\newcommand\boehm{\textsc{BDWGC}\xspace}

\begin{document}

\date{}  % This removes the date
\title{Response to Reviewer Feedback}
\maketitle

We thank the reviewers for their thoughtful reviews: we hope to have
taken onboard your feedback and addressed your questions. In this document
we summarise the major changes we have made: the diff shows changes to the prose
(though it does tend to mangle changes to figures and tables).

\subsection*{Finalization semantics}

\begin{blockquote}
explicitly discuss and present the issues around finalizations semantics, as
  learned from other languages
\end{blockquote}

We have made a number of changes (chiefly to the end of Section 3, and
throughout Section 4), which we hope address this.

As the reviewers note ``an
implementation that does not finalize anything would be compliant when
finalization may be deferred indefinitely'' is true of most GCed
implementations we are aware of, including Java. Indeed, Java once had the
\lstinline{runFinalizersOnExit} flag to force finalizers to be run before the
JVM exists, but it is marked as deprecated as ``inherently unsafe''; instead,
there is now the \lstinline{addShutdownHook} mechanism.

While investigating thus further, we discovered one surprise (to us, at least),
which muddies the waters with respect to finalizer latency further. In Rust,
when a thread unwinds (i.e.~panics) then, depending on compile-time flags, no
destructors will be run --- but Alloy can run all of the `queued' finalizers!
We have not belaboured this point in the paper, but it shows that destructors
can have guaranteed-infinite latency in some situations where finalizers have
infinite-at-worst latency.

@@

\subsection*{Performance Evaluation}

We have made a large number of changes to the performance evaluation. These are
mostly due to reviewer comments, but we also uncovered two mistakes in our
experiment: (1) we used metric XYZ instead of ABC in \boehm (2) we incorrectly
calculated confidence intervals in \laurie{which cases?}. While neither mistake
fundamentally changes the outcome, we believe it is important that we not
try and sweep this under the carpet.

The most significant changes are:

\begin{itemize}
  \item We provide a comparison of \lstinline{Rc<T>} with jemalloc and \boehm
    in Table~\laurie{3?} in the Appendix. This suggests that the inherent
    overhead of \boehm ranges from 2--26\%, averaging around 11\%.

  \item We have standardized all visualizations by normalizing against
    appropriate baselines (e.g.~Alloy-without-elision as the baseline for
    the \Eelision experiment; and barrier-free execution as the baseline
    for the \Epremopt experiment). This allows us to show what we hope
    are easily understand performance ratios. We also provide the raw data
    in the appendix.

  \item We provide data for wall-clock and user-time.
\end{itemize}

The downside to this is that we now present far too many figures and tables to
fit in the main body of the paper! We have tried to select the subset of
visualizations that will most quickly help the reader understand the overall
picture for the main body of the paper, relegating everything else to the
appendix.

We were asked -- when explaining the overhead of \boehm -- the following:

\begin{blockquote}
Ideally this would include a
discussion of how much of the performance deficit is likely redeemable via
"engineering", and how much is likely to be intrinsic (i.e. due to
unavoidable conservatism), so that we can get a better intuition of the
plausible "best case" for this approach.
\end{blockquote}

Although we can now quantify the overhead of Boehm vs.~jemalloc in our context,
the data doesn't really give us clues as to why the overhead exists. That makes
us nervous about speculating further. While we are not total allocator dummies,
both of us have talked to allocator experts over the years to know that there is a lot
here that we don't know. At a high level, \boehm is -- and we don't mean this as an
insult or criticism! -- ``old''. Since then many allocators (e.g.~dlmalloc, jemalloc)
have been designed, and subsequently tweaked and improved; each such allocator
is full of many decisions which are likely to affect its performance. Unfortunately,
we don't really know which decisions might be relevant or not.

@@

\subsection*{Conclusion}

\begin{blockquote}
add a discussion thats sums up the overall benefits/drawbacks of this
particular approach, possibly depending on specific scenarios/use cases, and
perhaps outlines directions for future work from a conceptional perspective,
i.e., possible design directions that could be investigated
\end{blockquote}

We have expanded the conclusions, but we hope to have done so in a way
that does not overstate our confidence in our predictions.

### 2025-07-28 response.tex

\begin{blockquote}
explicitly discuss and present the issues around finalizations semantics, as
  learned from other languages
\end{blockquote}

We have made a number of changes (chiefly to the end of Section 3, and
throughout Section 4), which we hope address this.

As the reviewers note ``an
implementation that does not finalize anything would be compliant when
finalization may be deferred indefinitely'' is true of most GCed
implementations we are aware of, including Java. Indeed, Java once had the
\lstinline{runFinalizersOnExit} flag to force finalizers to be run before the
JVM exists, but it is marked as deprecated as ``inherently unsafe''; instead,
there is now the \lstinline{addShutdownHook} mechanism.

While investigating this further, we discovered one surprise (to us, at least),
which muddies the waters with respect to finalizer latency further. In Rust,
when a thread unwinds (i.e.~panics) then, depending on compile-time flags, no
destructors will be run --- but Alloy can run all of the `queued' finalizers!
We have not belaboured this point in the paper, but it shows that destructors
can have guaranteed-infinite latency in some situations where finalizers have
infinite-at-worst latency.

@@

\subsection*{Performance Evaluation}

\jake{Since the original submission, alloy has been updated. We now
dynamically link boehm in all cases. For memory metrics, this alone won't
change things because we always used dynamic linkage for these tests anyway.
However, this may have affected wall-clock performance on some individual
benchmarks (though it's really hard to tell which, because there are
confounders: we also now run benchmarks on b15; we made changes to finalization
based on Andy wingo's feedback (mentioned below); and we made a couple of other
unrelated QoL and implementation changes to Alloy (i doubt these made a
difference, but who knows). What is clear though is that even if individual
benchmarks may have changed, the overall picture has not not since original
submission.}

We have made a large number of changes to the performance evaluation. These are
mostly due to reviewer comments, but we also uncovered a mistake in our
experiment. We used \boehm's \lstinline{GC_heapsize} metric (which never
shrinks) for calculating in-use heap snapshots, we have since modified boehm to
output the number of bytes swept and explicitly freed on each GC. While neither mistake
fundamentally changes the outcome, we believe it is important that we not
try and sweep this under the carpet. \jake{This does fundamentally change the
memory results for finalizer elision. In som-rs-ast, som-rs-bc, grmtools, and
ripgrep, finalizer elision now uses less memory than the no-elision
configuration. In the original submission, we incorrectly observed that it had
a higher average heap usage as a trade-off for better wall-clock performance.}
\jake{furthermore, since the original submission we have changed the way that
finalization works: we now use Boehm's \lstinline{finalize_on_demand} API based on
Andy Wingo's feedback. It is very possible that this could have moved the dial
on heap size too, how much that this changes things in relation to the above
mentioned wrong metric error I don't know. but it has crossed my mind.}

@@

The most significant changes are:

\begin{itemize}
  \item We provide a comparison of \lstinline{Rc<T>} with jemalloc and \boehm
    in Table~\laurie{3?} in the Appendix. This suggests that the inherent
    overhead of \boehm ranges from 2--26\%, averaging around 11\%.

  \item We have standardized all visualizations by normalizing against
    appropriate baselines (e.g.~Alloy-without-elision as the baseline for
    the \Eelision experiment; and barrier-free execution as the baseline
    for the \Epremopt experiment). This allows us to show what we hope
    are easily understand performance ratios. We also provide the raw data
    in the appendix.

  \item We provide data for wall-clock and user-time.

  \item We provide data for the number of collections run, and the total GC
  pause time.
\end{itemize}

@@

We were asked -- when explaining the overhead of \boehm -- the following:

\begin{blockquote}
Ideally this would include a
discussion of how much of the performance deficit is likely redeemable via
"engineering", and how much is likely to be intrinsic (i.e. due to
unavoidable conservatism), so that we can get a better intuition of the
plausible "best case" for this approach.
\end{blockquote}

Although we can now quantify the overhead of Boehm vs.~jemalloc in our context,
the data doesn't really give us clues as to why the overhead exists. That makes
us nervous about speculating further. While we are not total allocator dummies,
both of us have talked to allocator experts over the years to know that there is a lot
here that we don't know. At a high level, \boehm is -- and we don't mean this as an
insult or criticism! -- ``old''. Since then many allocators (e.g.~dlmalloc, jemalloc)
have been designed, and subsequently tweaked and improved; each such allocator
is full of many decisions which are likely to affect its performance. Unfortunately,
we don't really know which decisions might be relevant or not.

\jake{My reading of this reviewer comment is that our mention of allocator
differences are just a partial (yet important!) answer. I think they are also
asking something else too (based on their mention of ``unavoidable
conservatism''). This is more along the lines of: ``what GC-specific
engineering would be needed to make this research-based GC productionised to
have comparable performance with jemalloc RC''. My intuition is that they're
wondering about things like ``what if you had a Bartlett-style
semi-conservative collector rather than fully conservative?'' The short answer
here would also be ``it's too difficult to say'' IMHO. Bartlett-style
collection muddies the waters a bit showing that (counter to what I would
assume) precise object scanning doesn't make a lot of difference in terms of
wall-clock time. For some workloads its slower, for others its a modest
improvement. And trying to make guesses on Rust workloads based off of this
would be pure speculation on my part. Precise GC OTOH would be an interesting
comparison, maybe fully moving compaction would improve things... but as we
mention in the paper, this is not possible while respecting the Rust semantics
of today. If you think this is worth mentioning in the body of the paper I can
do so and include a citation to the Bartlett work. I don't know how on-topic it
is though.}

### 2025-07-28 response.tex

\subsection*{Finalization semantics}

\begin{blockquote}
explicitly discuss and present the issues around finalizations semantics, as
  learned from other languages
\end{blockquote}

We have made a number of changes (chiefly to the end of Section 3, and
throughout Section 4), which we hope address this.

As the reviewers note ``an
implementation that does not finalize anything would be compliant when
finalization may be deferred indefinitely'' is true of most GCed
implementations we are aware of, including Java. Indeed, Java once had the
\lstinline{runFinalizersOnExit} flag to force finalizers to be run before the
JVM exists; it was long marked as deprecated as ``inherently unsafe'' and
now appears to have been removed; instead,
there is now the \lstinline{addShutdownHook} mechanism.

While investigating this further, we discovered one surprise (to us, at least),
which muddies the waters with respect to finalizer latency further. In Rust,
when a thread unwinds (i.e.~panics) then, depending on compile-time flags, no
destructors will be run --- but Alloy can run all of the `queued' finalizers!
We have not belaboured this point in the paper, but it shows that destructors
can have guaranteed-infinite latency in some situations where finalizers have
infinite-at-worst latency.

@@

\subsection*{Performance Evaluation}

We have made a large number of changes to the performance evaluation. These are
mostly due to reviewer comments, but we also uncovered a mistake in our
experiment. We previously used \boehm's \lstinline{GC_heapsize} metric:
this does not, as we assumed, represent current heap usage, but total
memory usage since the program's start. We now use the \laurie{XYZ} metric,
which does measure the correct thing. This changes the results for finalizer
elision in on way: previously this appeared to cause heap usage to grow, but in
fact it shrinks.

@@

The most significant changes are:

\begin{itemize}
  \item We provide a comparison of \lstinline{Rc<T>} with jemalloc and \boehm
    in Table~\laurie{3?} in the Appendix. This suggests that the inherent
    overhead of \boehm ranges from 2--26\%, averaging around 11\%.

  \item We have standardized all visualizations by normalizing against
    appropriate baselines (e.g.~Alloy-without-elision as the baseline for
    the \Eelision experiment; and barrier-free execution as the baseline
    for the \Epremopt experiment). This allows us to show what we hope
    are easily understand performance ratios. We also provide the raw data
    in the appendix.

  \item We provide data for wall-clock and user-time.

  \item We provide data for the number of collections run, and the total GC
  pause time.
\end{itemize}

We also made some minor changes (e.g.~we now always use dynamic
linking of \boehm; we reran benchmarks on a more modern machine), but
these had no obvious impact on our results.

@@

The downside to this is that we now present far too many figures and tables to
fit in the main body of the paper! We have tried to select the subset of
visualizations that will most quickly help the reader understand the overall
picture for the main body of the paper, relegating everything else to the
appendix.

We were asked -- when explaining the overhead of \boehm -- the following:

\begin{blockquote}
Ideally this would include a
discussion of how much of the performance deficit is likely redeemable via
"engineering", and how much is likely to be intrinsic (i.e. due to
unavoidable conservatism), so that we can get a better intuition of the
plausible "best case" for this approach.
\end{blockquote}

We ended up a little unsure which of two questions this might be asking,
so please forgive us if we answer both!

@@

Although we can now quantify the overhead of Boehm vs.~jemalloc in our context,
the data doesn't really give us clues as to why the overhead exists. That makes
us nervous about speculating further. While we are not total allocator dummies,
both of us have talked to allocator experts over the years to know that there is a lot
here that we don't know. At a high level, \boehm is -- and we don't mean this as an
insult or criticism! -- ``old''. Since then many allocators (e.g.~dlmalloc, jemalloc)
have been designed, and subsequently tweaked and improved; each such allocator
is full of many decisions which are likely to affect its performance. Unfortunately,
we don't really know which decisions might be relevant or not.

Another possibility is to use a Bartlett-esque semi-conservative GC. The literature
offers a complex picture: sometimes this is faster, sometimes slower, than
fully conservative GC. We again lack confidence to know how this might translate
into our context.

### 2025-07-29 tables/exclusions.tex
Update benchmark exclusions table in appendix

\begin{tabular}{lll}
\toprule
Benchmark         & Description                              & Reason for exclusion                                       \\
\midrule
bevy          & ECS game engine in Rust                   & \jake{TODO} \\
dyon          & Scripting language in Rust                & \jake{TODO} \\
jiff          & A datetime library for Rust               & \jake{TODO} \\
mini-moka     & Concurrent in-memory cache library        & \jake{TODO} \\
numbat        & Math search engine                        & \jake{TODO} \\
rkyv          & Zero-copy deserialization framework       & \jake{TODO} \\
RustPython    & Python interpreter written in Rust        & \jake{TODO} \\
rust-analyzer & Language server for Rust                  & \jake{TODO} \\
salsa         & Incremental recomputation library         & \jake{TODO} \\
WLambda       & scripting language in Rust                & \jake{TODO} \\
\bottomrule
\end{tabular}

>>>>

\begin{tabular}{lll}
\toprule
Benchmark         & Description                              & Reason for exclusion                                       \\
\midrule
bevy          & ECS game engine in Rust                   & Unable to port successfully (see \cref{rustpython}) \\
dyon          & Scripting language in Rust                & Unable to port successfully (see \cref{rustpython})\\
jiff          & A datetime library for Rust               & Too few allocations to measure \\
mini-moka     & Concurrent in-memory cache library        & Too few allocations to measure \\
numbat        & Math search engine                        & Too few allocations to measure \\
rkyv          & Zero-copy deserialization framework       & Insufficient \texttt{Gc<T>} coverage in benchmarks \\
RustPython    & Python interpreter written in Rust        & Difficulty retro-fitting \texttt{\_\_del\_\_} semantics (see \cref{rustpython}). \\
rust-analyzer & Language server for Rust                  & Unable to port successfully (see \cref{rustpython}) \\
salsa         & Incremental recomputation library         & Too few allocations to measure \\
WLambda       & Scripting language written in Rust        & Insufficient \texttt{Gc<T>} coverage in benchmarks \\
\bottomrule
\end{tabular}
